---
title: "cobiclust: _Erysiphe alphitoides_ pathobiome tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cobiclust: _Erysiphe alphitoides_ pathobiome tutorial}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Introduction

This tutorial intends you to use  `cobiclust` to find structure in the *_Erysiphe alphitoides_ pathobiome dataset* kindly provided by [Corinne Vacher](https://corinnevacher.wordpress.com/people/).

It consists of metabarcoding data of leaf-associated microbiome of 114 leaves sampled from 3 different oaks, with different status with respect to a pathogen. Two barcodes were used: the 16S for the bacterial fraction and the ITS1 for the fungal fraction. One taxa (*Erysiphe alphitoides*) is of particular interest as it is the causal agent of oak powdery mildew. The goal of the original study ([10.1007/s00248-016-0777-x](https://link.springer.com/article/10.1007/s00248-016-0777-x)) was to assess the impact the impact of *Erysiphe alphitoides* on the foliar fungal and bacterial communities.

Details on the sampling protocol and bioinformatic analyses of the raw reads are available in the original [publication](https://link.springer.com/article/10.1007/s00248-016-0777-x). 

The data have been preprocessed and filtered (to remove low-abundance, low abundance taxa). They are available from the website

- [count data](https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/Data/counts.tsv)
- [metadata](https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/Data/metadata.tsv)
<!-- - [offsets](https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/Data/offsets.tsv) -->

[Details](https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/PLN_oaks.html) on the metadata are available from the website.

# Loading useful packages or functions

We first load `cobiclust` (`ggplot2` for some graphics, `purr` and `dplyr` for some data manipulations) and `pathobiome_usefulfun.R` providing functions useful for this example.

```{r setup, cache = FALSE}
library(cobiclust)
library(ggplot2)
library(purrr)
library(dplyr)
```

# Importing the data

We then import the data in R and format them for `cobiclust`. The fungal and bacterial communities were sequenced using different barcodes and should thus have different offset. We compute offsets matrix with the Total Counts method using the function `calculate_offset_perkingdom`.

```{r, echo = FALSE}
calculate_offset_perkingdom <- function(counts = counts) {
  kingdom <-  unlist(lapply(strsplit(rownames(counts), split = "_"), FUN = function(y) y[[1]]))
  ou_fungi <- which(kingdom %in% c("f","E"))
  TC_fungi <- colSums(counts[ou_fungi,])/mean(colSums(counts[ou_fungi,])) 
  ou_b <- which(kingdom=="b")
  TC_bact <- colSums(counts[ou_b,])/mean(colSums(counts[ou_b,])) 
  
  # Matrix of nu_j
  indic <- c(rep(1, length(which(kingdom %in% c("f","E")))), rep(0,length(which(kingdom == "b"))))
  indic2 <- as.matrix(cbind(indic, abs(indic-1)))
  mat_nuj <- indic2%*%matrix(ncol = ncol(counts), nrow = 2, c(TC_fungi, TC_bact), byrow = TRUE)
  return(mat_nuj)
}
```


```{r}
counts <- read.table(file = "https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/Data/counts.tsv")
metadata <- read.table(file = "https://mia-paris.pages.mia.inra.fr/formation_abondance_reseau/tutoriels/PLN_TP/Data/metadata.tsv")
```
# Preparing the data

To be on the safe side, let's transform `counts` and `offsets` to matrices (they are imported as `data.frame` by default), and transpose then to have the OTUs in rows and the leaves in columns.

```{r}
counts <- as(t(counts), "matrix")
# Calculate offsets matrix
mat_nuj <- calculate_offset_perkingdom(counts)
```

Let's calculate the percentage of zero counts.
```{r}
length(which(counts==0))*100/length(counts)
```

# Fitting several Poisson-Gamma Latent Block Model

We here fit the Poisson-Gamma Latent Block Model implemented in the `cobiclust` package with a block-specific dispersion parameter (`akg = TRUE`) for $K = 1, ..., 8$ and $G = 1, ..., 6$.

Be caution : this chunk may be quite long to terminate; so we to not evaluate here.

```{r cobiclust_akg, eval = FALSE}
Kmax <- 8
Gmax <- 6
arg2 <- rep(rep(1:Kmax), Gmax)
arg3 <- rep(1:Gmax, each = Kmax)

res_akg <-  pmap(list(arg2, arg3), ~ cobiclust(counts, ..1, ..2, akg = TRUE, nu_j = mat_nuj))
```

# Selection of the best Poisson-Gamma Latent Block Model

```{r best_model, eval = FALSE}
arg1 <- 1:(Kmax*Gmax)
# Calculation of the variational BIC and variational ICL criteria
crit_akg <- pmap(list(arg1, arg2, arg3), ~ selection_criteria(res_akg[[..1]], K = ..2, G = ..3), )
crit_akg <- data.frame(Reduce(rbind, crit_akg)) 
# Selection of the best model
vICL <- crit_akg %>% dplyr::filter(vICL == max(vICL)) 
vICL
ou_best <- apply( vICL[,c(1,6:7)], 1, FUN=function(x) intersect(which(arg2==x[2]),which(arg3==x[3])))
best_akg <- res_akg[[ou_best]]
```
```{r}
# We directly fit the best model for this example
best_akg <- cobiclust(counts, K = 5, G = 3, akg = TRUE, nu_j = mat_nuj)
```


# Some informations   

## Groups in rows and columns
```{r}
# Groups in rows
Z <- best_akg$classification$rowclass
# Groups in columns
W <- best_akg$classification$colclass
table(Z)
table(W)
# Description of the groups
#sapply(1:max(W),FUN=function(x) table(colnames(counts)[W==x]))
sapply(1:max(Z),FUN=function(x) rownames(counts)[Z==x])
table(data.frame(metadata$tree,W))
```
## Models parameters
```{r info}
# alpha_kg
print(best_akg$parameters$alpha)
# a
print(best_akg$parameters$a)
```

## Some useful graphics

### Shannon diversity Index according to $W$
Let's represent the Shannon diversity Index, a common measure of diversity using the proportional abundances of each species, calculated for each leaf, according to the group of leaves.

```{r graph1}
don <- counts
p_i <- apply(don, 2, FUN = function(x) x / sum(x))
H <- apply(p_i, 2, FUN = function(x) - sum(x * log2(x), na.rm = TRUE))
df <- data.frame(Leaf = 1:ncol(don), H = H, W = factor(round(best_akg$info$t_jg %*% (1:max(W)))))
levels(df$W) <- paste("W",levels(df$W),sep="")
hwplot <- ggplot(data = df, aes(x = W, y = H))
hwplot <- hwplot + geom_boxplot() + labs(x = "")
print(hwplot)
rm(p_i, df)
```

### Data reordered according to the biclustering
```{r graph2}
image(log(counts+0.5), main = "Raw data on log-scale")
image(log(counts[order(Z), order(W)]+0.5), main = "Reordered data on log-scale")
```

### Boxplot of level of infection in log-scale for the groups of leaves

```{r graph3}
plot.infection <- ggplot(data = data.frame(pmInfection = log2(metadata$pmInfection), W = as.factor(paste("W",W,sep=""))), aes(y = pmInfection, x = W))
plot.infection <- plot.infection + geom_boxplot() + labs(x = "", y = "Level of infection") + theme_bw()
print(plot.infection)
```

### Boxplot of the $\hat{\mu_i}$ within each group in row $Z_k$

```{r graph4}
df1 <- data.frame(Microorg = rownames(counts), Z = round(best_akg$info$s_ik %*% (1:max(Z))))
df2 <- data.frame(Microorg = rownames(counts), Mu = best_akg$parameters$mu_i)
df <- merge(df1, df2, by = "Microorg")
tmp <- sapply(1:max(Z), FUN = function(x) mean(best_akg$parameters$mu_i[Z == x]))
df$Z <- factor(df$Z)
levels(df$Z) <- paste("Z", levels(df$Z), sep = "")
muzplot <- ggplot(data = df, aes(x = Z, y = log(Mu)))
muzplot <- muzplot + geom_boxplot() + labs(x = "Z") 
muzplot <- muzplot + theme_bw() +  labs(x = "", y = expression(mu[i])) 
print(muzplot)
rm(df1, df2, df, tmp)
```

### Plot of the vICL criterion

We propose to plot the $vICL$ criterion according to the (y-axis) to the number of groups in rows $K$ (x-axis). Colours depends on $G$ values.
```{r graph5, eval = FALSE}
crit_akg$G <- as.factor(crit_akg$G)
plot.vICL_K <- ggplot(data = crit_akg, aes(y = vICL, x = K, group = G, colour = G))

plot.vICL_K <- plot.vICL_K + geom_point()  + geom_line() +  theme_bw() + theme(legend.position = "bottom") 
print(plot.vICL_K)
```

# Remarks

This vignette is designed to help you work with `cobiclust`. Naturally, you can use your imagination to represent or visualize the results.

I would like to thank Mahendra Mariadassou, Julien Chiquet and Stéphane Robin who wrote a vignette on the use of [`PLNModels`](http://julien.cremeriefamily.info/PLNmodels/) based on the same example and from which I was very strongly inspired for the beginning of this vignette.


